Functors, CPS and monads, or how to generate efficient code from abstract designs

### 2005-04-03 paper/metamonads.tex
part of section 4 written.  slow going...

>>>>

Now that we know that we have the appropriate basic techniques for
writing our code generators, we need to worry about modularizing them
appropriately.  Here we can really take advantage of the fact that
MetaOCaml is a staged language: it is not really relevant how much
computation happens at staging time, as long as the generated code
is correct and efficient.  In other words, we can use any abstraction
mechanisms we want to structure our code generators, as long as none
of those abstractions infiltrate the generated code.

While the Object-Oriented Design community has acquired an extensive
vocabulary for describing modularity ideas, the guiding principles for
modular designs has not changed since they were first articulated by
Parnas~\cite{journals/cacm/parnas72a} and Dijkstra~
\cite{EWD:EWD447}: information hiding and separation of concerns.  To
apply these principles to the study of Gaussian Elimination, we need
to understand what changes between different implementations, and 
what concerns need to be addressed.  We also need to study the degree
to which these concerns are independent.

A careful study of Gaussian Elimination \cite{carette04} shows that
(at least) the following variations occur:
\begin{enumerate}
	\item \textbf{Domain}: In which (algebraic) domain do the matrix
		elements belong to.  Some implementations were very specific
		($\mathbb{Z}, \mathbb{Q}, \mathbb{Z_p}, 
		\mathbb{Z}_p\left[\alpha_1,\ldots,\alpha_n\right], 
		\mathbb{Z}\left[x\right]$, $\mathbb{Q}\left(x\right)$, 
		$\mathbb{Q}\left[\alpha\right]$, and floating point numbers for 
		example), while others were generic for elements of a field,
		multivariate polynomials over a field, elements of a formal Ring
		with possibly undecidable zero-equivalence, or elements of a 
		Division Ring.  In the roughly 85 pieces of code surveyed,
		20 different domains were encountered.
	\item \textbf{Container}: Whether the matrix
		was represented as an array of arrays, a one-dimensional array,
		a hash table, a sparse matrix, etc., and
		whether indexing was done in C or Fortran style.  Additionally,
		if a particular representation had a special mechanism for efficient
		row exchanges, this was sometimes used.  
	\item \textbf{Output choices}:  Whether just the reduced matrix, or
		additionally the rank, the determinant, and the
		pivoting matrix were to be returned.
		It is worthwhile noting that in the larger algorithm family,
		routines like Maple's \texttt{LinearAlgebra:-LUDecomposition} have
		up to $2^6 + 2^5 + 2^2 = 100$ possible outputs.
	\item \textbf{Fraction-free}: Whether the Gaussian Elimination
		algorithm is allowed to use unrestricted division, or only
		exact (remainder-free) division.
	\item \textbf{Pivoting}: Whether to use no pivoting at all, only
		column-wise pivoting, or full pivoting.
	\item \textbf{Augmented Matrices}: Whether we are doing GE on
		a full matrix, or only a restricted number of columns, while
		doing elimination on the full matrix.  Note that we currently
		do not treat this aspect in our code.
\end{enumerate}
\noindent In addition to the above variations, there are two aspects that 
recur frequently:
\begin{enumerate}
	\item \textbf{Length measure}:  For stability reasons
		(whether numerical or coefficient growth), if a domain possesses
		an appropriate length measure, this was sometimes used to choose
		an ``optimal'' pivot for row exchanges.
	\item \textbf{Normalization and zero-equivalence}: Whether the 
		arithmetic operations of the domain at hand gives results in 
		normalized form, and whether a specialized zero-equivalence 
		routine needs to be used.
\end{enumerate}
\noindent These are separated out from the others are they are cross-cutting
concerns: in the case of the length measure, then a property of the domain
will influence the method used to do pivoting \textbf{if} pivoting is to be
performed.
